---
title: "Reproducible Research - Assessment 1"
author: "Sambasiva Andaluri"
date: "April 9, 2015"
output: html_document
---

### Load packages

```{r,message=FALSE}
library(plyr)
library(dplyr)
library(lubridate)
library(lattice)
```

### Loading and preprocessing the data

**1. Load the data.**


```{r}
rawActivityData <- read.csv("activity.csv")
```


**2. Process/transform the data (if necessary) into a format suitable for your analysis**

```{r}
activityNoNA <- rawActivityData[!is.na(rawActivityData),]
```

### What is mean total number of steps taken per day?

**1. Calculate the total number of steps taken per day**

```{r}
stepsTotalByDay <- activityNoNA %>% group_by(date) %>% summarise(total=sum(steps))
stepsTotalByDay <- stepsTotalByDay[!is.na(stepsTotalByDay$total),]
knitr::kable(stepsTotalByDay, align=c('l', 'l'), col.names=c("Date", "Total Steps"))
```

**2. Make a histogram of the total number of steps taken each day**

```{r}
hist(stepsTotalByDay$total, col="blue", main="Histogram of Total daily steps", xlab="Total steps per day")
```

**3. Calculate and report the mean and median of the total number of steps taken per day.**

```{r}
mean(stepsTotalByDay$total)
median(stepsTotalByDay$total)
```

### What is the average daily activity pattern?

**1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**

```{r}
stepsAvgByInterval <- activityNoNA %>% group_by(interval) %>% summarise(average=round(mean(steps)))
stepsAvgByInterval <- stepsAvgByInterval[!is.na(stepsAvgByInterval$interval),]
plot(stepsAvgByInterval, type="l", xlab="Intervals (5 min each)", ylab="Average number of steps per interval")
```

**2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? **

```{r}
stepsAvgByInterval[which(stepsAvgByInterval$average==max(stepsAvgByInterval$average)),1]
```

### Imputing missing values

**1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)**

```{r}
nrow(rawActivityData[is.na(rawActivityData),])
```

**2. Devise a strategy for filling in all of the missing values in the dataset. **

**Strategy used: Using mean of steps by interval as value for imputing NAs in the original dataset**

```{r}
  activityDataImputed <- rawActivityData
  for (i in 1:nrow(activityDataImputed)) { 
      if( is.na(activityDataImputed$steps[i] )) { 
        activityDataImputed[i, "stepsImputed"] <- 
          round(stepsAvgByInterval[match(activityDataImputed$interval[i], stepsAvgByInterval$interval),2])
      } else {
         activityDataImputed[i, "stepsImputed"] <- activityDataImputed$steps[i]  
      }
  }
```

**3. Create a new dataset that is equal to the original dataset but with the missing data filled in.**

```{r}
nrow(activityDataImputed[is.na(activityDataImputed$stepsImputed),])
str(activityDataImputed)
```

**4. Make a histogram of the total number of steps taken each day.**

```{r}
stepsTotalByDayImputed <- activityDataImputed %>% group_by(date) %>% summarise(total=sum(stepsImputed))
stepsTotalByDayImputed <- stepsTotalByDayImputed[!is.na(stepsTotalByDayImputed$total),]
hist(stepsTotalByDayImputed$total, col="red", main="Histogram of Total daily steps", xlab="Total steps per day")
```

**4. Calculate and report the mean and median total number of steps taken per day.**

```{r}
mean(stepsTotalByDayImputed$total)
median(stepsTotalByDayImputed$total)
```

### Are there differences in activity patterns between weekdays and weekends?

**1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.**

```{r}
    for (i in 1:nrow(activityDataImputed)) { 
       wday <- wday(activityDataImputed$date[i], label=FALSE)
       if (wday > 1 && wday < 7) {
         activityDataImputed$weekday[i] <- "weekday"
       } else {
         activityDataImputed$weekday[i] <- "weekend"
       }
    }
    activityDataImputed$weekday <- factor(activityDataImputed$weekday)
    str(activityDataImputed)
```

**2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). **

```{r}
avgStepsByWeekdayIntervalImputed <- activityDataImputed %>% group_by(weekday, interval) %>% summarise(average=mean(stepsImputed))
xyplot(average ~ interval | weekday, data = avgStepsByWeekdayIntervalImputed, xlab="Interval", ylab="Steps", type="l", layout=c(1,2))
```

